---
title: "A brief tutorial on runing Maxent in R"
author: "Xiao Feng, Cassondra Walker, Fikirte Gebresenbet"
date: "July 4, 2017"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    keep_md: yes
    toc: yes
    toc_depth: 3
—dsadas
## 1. Set up the working environment  

### 1.1 Load packages  
Running Maxent in R requires several packages. Specifically, the "dismo" package, which contains the functions for maximum entropy species distribution modeling. However, the "rgbif" package gives access to the GBIF database for occurrence location downloads, (https://cran.r-project.org/web/packages/rgbif/rgbif.pdf), the "raster" package provides functions for working with gridded data ( https://cran.r-project.org/web/packages/raster/raster.pdf), the "rgeos" package provides the ability to manipulate spatial data (https://cran.rstudio.com/web/packages/rgeos/rgeos.pdf), and the "knitr" package allows for report generation and easy display of modeling output (https://cran.r-project.org/web/packages/knitr/knitr.pdf), which are needed to complete the modeling process.  

##### Thread 1  
```{r setup1,message=FALSE}
library("dismo")
library("raster")
library("rgbif")
library("knitr")
require("rgeos")
dyn.load('/Library/Java/JavaVirtualMachines/jdk1.8.0_144.jdk/Contents/Home/jre/lib/server/libjvm.dylib')# a special step for Mac system
require("rJava")
#knitr::opts_knit$set(root.dir = 'd:/projects/2017_7_workshop_enm_R')
knitr::opts_knit$set(root.dir = '/Users/iel82user/Google Drive/1_osu_lab/projects/2017_7_workshop_enm_R')
```

###1.2 Set up the Maxent path  
```{r maxent, eval=FALSE, echo=TRUE}
# download maxent.jar 3.3.3k, and place the file in the desired folder
utils::download.file(url="https://raw.githubusercontent.com/mrmaxent/Maxent/master/ArchivedReleases/3.3.3k/maxent.jar",destfile=paste0(system.file("java", package="dismo"),"/maxent.jar"),
                     mode="wb") ## wb for binary file, otherwise maxent.jar can not execute
# also note that both R and Java need to be the same bit (either 32 or 64) to be compatible to run

#options( java.parameters = c("-Xss2560k", "-Xmx2g") ) ## increase memory size of the JVM, this code may prevent memory issues of Maxent.jar

```

##2. Prepare data input  
###2.1 Load environmental layers  
```{r load rasters}
# load GIS layers; here we used bioclim variables as an example (downloaded from worldclim.org)
clim <- list.files("data/bioclim/",pattern=".bil$",full.names = T)
clim <- raster::stack(clim) ## stacking the bioclim variables to process them at one go
```

###2.2 Occurrence data  
###2.2.1 Download occurrence data  
```{r prepare, message=TRUE, warning=FALSE}
# download occurrence data from GBIF; for example, we used nine-banded armadillo that distributes in America
if(file.exists("data/occ_raw")){
  #cat(1)
  load("data/occ_raw")
}else{
  #cat(2)
  occ_raw <- gbif("Dasypus novemcinctus")
  save(occ_raw,file = "data/occ_raw")
  write.csv("data/occ_raw.csv")
}
#head(occ_raw)
```

###2.2.2 Clean occurrence data
```{r clean data}
# remove bad coordinates, where either the lat or long coordinate is missing
occ_clean <- subset(occ_raw,(!is.na(lat))&(!is.na(lon)))
cat(nrow(occ_raw)-nrow(occ_clean), "records are removed")

# remove duplicated data based on latitude and longitude
dups <- duplicated(occ_clean[c("lat","lon")])
occ_unique <- occ_clean[!dups,]
cat(nrow(occ_clean)-nrow(occ_unique), "records are removed")

# make occ spatial
coordinates(occ_unique) <- ~ lon + lat
plot(occ_unique) ## we may notice an erroneous point

# remove some errors
occ_unique <- occ_unique[which(occ_unique$lon>-110 &
                                 occ_unique$lon < -40),]

# make occ sparse (keep one occ per cell)
cells <- cellFromXY(clim[[1]],occ_unique)
dups <- duplicated(cells)
occ_final <- occ_unique[!dups,]
cat(nrow(occ_unique)-nrow(occ_final), "records are removed")

plot(clim[[1]]) ## to draw the first layer (or replace [[1]] with any nth number of the layers with in the raster stack)
plot(occ_final,add=T,col="red") ## the 'add=T' tells R to put the incoming data on the existing layer (the clim in this case)
```

###2.3 Set up study area
```{r set up study area}
# this creates a buffer around the occurence data 
occ_buff <- buffer(occ_final,4) ## 4 decimal degree
plot(clim[[1]]) ## this plots the first element ([[1]]) in the raster stack and adds the occurence data
plot(occ_final,add=T,col="red") ## this adds the occurrence data
plot(occ_buff,add=T,col="blue") ## this adds the buffer polygon

# if the area we will mask from is very large, use crop first; it will crop from a rectangle and then mask - that will save a lot of time. 
studyArea <- crop(clim,extent(occ_buff))  ## gives a coarser rectangle of the study area (a rectangle encompassing the buffered area). 

studyArea <- mask(studyArea,occ_buff) ## the 'study area' created by cropping and masking the raste stacks is also going to remain a raster stack 

# save the buffer areas into raster files
writeRaster(studyArea,
            filename=paste0("data/studyarea/",names(studyArea),".asc"), ## a series of names for output files
            format="ascii", ## the output format
            bylayer=TRUE, ## this will save a series of layers
            overwrite=T)

# select background points from this buffered area
set.seed(1) ## when the number provided to set.seed() function, the same random sample will be selected in the next line; use this code before the sampleRandom function everytime, if you want to ge the same "random samples"

bg <- sampleRandom(x=studyArea,
                   size=10000,
                   na.rm=T, ## na.rm is telling R to remove the 'Not Applicaple' points, 
                   sp=T) ## sp is telling R to give us a spatially points 
plot(studyArea[[1]])
plot(bg,add=T) ## add the background points to the plotted raster
plot(occ_final,add=T,col="red") ## add the occurrence data to the plottted raster

```

###2.4 Split occurrence data into training & testing
```{r cut occ into training & testing}
# randomly select 50% for training
set.seed(1) ## get the same random sample for training and testing
selected <- sample(1:nrow(occ_final),nrow(occ_final)*0.5)
occ_train <- occ_final[selected,] ## this is the selection
occ_test <- occ_final[-selected,] ## this is the opposite of the selection
```

###2.5 Format data for Maxent
```{r prepare data for maxent,message=FALSE}
# extracting env conditions for training occ from the raster stack; a data frame is returned 
p <- extract(clim,occ_train) ## env conditions for training occ this makes a dataframe since environmental conditions extracted are from the raster stack (i.e multiple columns)
p_test <- extract(clim,occ_test) ## env conditions for testing occ
a <- extract(clim,bg)  ## env conditions for background
pa <- c(rep(1,nrow(p)), rep(0,nrow(a))) ## repeat the number 1 as many numbers as the number of raws in my p
## (rep(1,nrow(p)) creating the number of rows as the p data set to have the number one as the indicator for presence
## rep(0,nrow(a)) creating the number of rows as the a data set to have the number zero as the indicator for absence
## The c combines these ones and zeros into a new vector that can be added to the Maxent table
pder <- as.data.frame(rbind(p,a)) ## this makes a data frame with the environmental attributes of the presence and absence locations
```

##3 Maxent models
###3.1 Simple implementation
```{r simple maxent model}
# mod <- maxent(x=clim, p=occ_train)
mod <- maxent(x=pder, ## env conditions
              p=pa,   ## 1:presence or 0:absence
              path=paste0(getwd(),"/output/maxent_outputs"), ## folder to store maxent output; if we do not specify a folder R will put the results in a temp file, and it gets messy to read those. . .
              args=c("responsecurves") ## a lot of parameters can be specified here
              )
## the maxent functions runs a model in the default settings..to change these parameters, you have to tell it what you want...i.e. response curves or the the type of features

# view a maxent model in a html brower
mod

# view detailed results
mod@results
```

###3.2 Predict function
```{r predict}
# maxent.R doesnt give us a prediction of training data/layers (unless you specify the projection layers in the "args""), the alternative is to use the predict function 
# a maxent model (in R) can be projected on raster layers or a dataframes

# example 1, project to out study area [raster]
ped1 <- predict(mod,studyArea)
plot(ped1)

# example 2, project to the world
#ped2 <- predict(mod,clim)
#plot(ped2)

# example 3, project training occurrences [dataframes]
ped3 <- predict(mod,p)
head(ped3)
hist(ped3)
```

###3.3 Model evaluation
```{r model evaluation}
# using "training data" to evaluate 
# "evaluate"" is an evaluation function from dismo package; p= presence and a=background
mod_eval_train <- dismo::evaluate(p=p,a=a,model=mod) #p & a are dataframes (the p and a are the training presence and training absence points, and vice verse in the testing as well (below))
print(mod_eval_train)
# This is the test AUC
mod_eval_test <- dismo::evaluate(p=p_test,a=a,model=mod)  
print(mod_eval_test) # training AUC may be higher than testing AUC

# calculate thresholds of models
# threshold function is in dismo and based on the evaluation function
thd1 <- threshold(mod_eval_train,"no_omission") # 0% omission rate [minimum training presence]
thd2 <- threshold(mod_eval_train,"spec_sens") # hiest TSS

# Only plot the predictions that are higher than the threshold....giving a binary output
plot(ped1>=thd1)  ## plotting points that are above the previously calculated tresholded value 
```

##4 Maxent parameters
###4.1 Select features
```{r detailed parameters}
# load the function that prepares parameters for maxent
source("code/Appendix2_prepPara.R")

mod1_autofeature <- maxent(x=pder[c("bio1","bio4","bio11")], ## env conditions, here we selected only 3 predictors
               p=pa, ## 1:presence or 0:absence
              path=paste0(getwd(),"/output/maxent_outputs1_auto"), ## path of maxent output, this is the folder you will find manxent output
              args=prepPara(userfeatures=NULL) )  ## default is autofeature
              
# or select Linear& Quadratic features
mod1_lq <- maxent(x=pder[c("bio1","bio4","bio11")], ## env conditions, here we selected only 3 predictors
               p=pa, ## 1:presence or 0:absence
              path=paste0(getwd(),"/output/maxent_outputs1_lq"), ## path of maxent output, this is the folder you will find manxent output
              args=prepPara(userfeatures="LQ") )  ## default is autofeature, here LQ represents Linear& Quadratic (L-linear, Q-Quadratic, H-Hinge, P-Product, T-Threshold)
        
```

###4.2 Change beta-multiplier
```{r beta-multiplier}
#change betamultiplier for all features
mod2 <- maxent(x=pder[c("bio1","bio4","bio11")], 
               p=pa, 
              path=paste0(getwd(),"/output/maxent_outputs2_o.5"), 
              args=prepPara(userfeatures="LQ",
                            betamultiplier=0.5) ) 

mod2 <- maxent(x=pder[c("bio1","bio4","bio11")], 
               p=pa, 
              path=paste0(getwd(),"/output/maxent_outputs2_complex"), 
              args=prepPara(userfeatures="LQH", ## include L, Q, H features
                            beta_lqp=1.5, ## use different betamultiplier for different features
                            beta_hinge=0.5 ) ) 
```

###4.3 Specify projection layers
```{r specify projection layers/data table}
# note: 1)the projection layers must exist in the hard disk (as relative to computer RAM); (2) the names of the layers (excluding the name extension) must match the names of the predictor variables; 
mod3 <- maxent(x=pder[c("bio1","bio11")], 
               p=pa, 
              path=paste0(getwd(),"/output/maxent_outputs3_prj1"), 
              args=prepPara(userfeatures="LQ",
                            betamultiplier=1,                      projectionlayers="/Users/iel82user/Google Drive/1_osu_lab/projects/2017_7_workshop_enm_R/data/studyarea") ) 

# load the projected map
ped <- raster(paste0(getwd(),"/output/maxent_outputs3_prj1/species_studyarea.asc"))
plot(ped)

# we can also project on a broader map, but please caustion about the inaccuracy associated with model extrapolation.
mod3 <- maxent(x=pder[c("bio1","bio11")], 
               p=pa, 
              path=paste0(getwd(),"/output/maxent_outputs3_prj2"), 
              args=prepPara(userfeatures="LQ",
                            betamultiplier=1,                      projectionlayers="/Users/iel82user/Google Drive/1_osu_lab/projects/2017_7_workshop_enm_R/data/bioclim") ) 
# plot the map
ped <- raster(paste0(getwd(),"/output/maxent_outputs3_prj2/species_bioclim.asc"))
plot(ped)

# simply check the difference if we used a different betamultiplier
mod3_beta1 <- maxent(x=pder[c("bio1","bio11")], 
               p=pa, 
              path=paste0(getwd(),"/output/maxent_outputs3_prj3"), 
              args=prepPara(userfeatures="LQ",
                            betamultiplier=100, ## for an extreme example, set beta as 100
                            projectionlayers="/Users/iel82user/Google Drive/1_osu_lab/projects/2017_7_workshop_enm_R/data/bioclim") ) 
ped3 <- raster(paste0(getwd(),"/output/maxent_outputs3_prj3/species_bioclim.asc"))
plot(ped-ped3) ## quickly check the difference between the two predictions

```

###4.4 Clamping function
```{r clamping function}
# enable or disable clamping function; note clamping function is involved when projecting
mod4_clamp <- maxent(x=pder[c("bio1","bio11")],
                     p=pa,
                     path=paste0(getwd(),"/output/maxent_outputs4_clamp"), 
                     args=prepPara(userfeatures="LQ",
                                   betamultiplier=1,
                                   doclamp = TRUE,
                                   projectionlayers="/Users/iel82user/Google Drive/1_osu_lab/projects/2017_7_workshop_enm_R/data/bioclim")) 

mod4_noclamp <- maxent(x=pder[c("bio1","bio11")], 
                       p=pa, 
                       path=paste0(getwd(),"/output/maxent_outputs4_noclamp"),
                       args=prepPara(userfeatures="LQ",
                                      betamultiplier=1,
                                      doclamp = FALSE,
                                      projectionlayers="/Users/iel82user/Google Drive/1_osu_lab/projects/2017_7_workshop_enm_R/data/bioclim") ) 


ped_clamp <- raster(paste0(getwd(),"/output/maxent_outputs4_clamp/species_bioclim.asc") )
ped_noclamp <- raster(paste0(getwd(),"/output/maxent_outputs4_noclamp/species_bioclim.asc") )
plot(stack(ped_clamp,ped_noclamp))
plot(ped_clamp - ped_noclamp) ## we may notice small difference, especially clamp shows higher predictions in most areas.


```

###4.5 Cross validation
```{r cross validation}
mod4_cross <- maxent(x=pder[c("bio1","bio11")], p=pa, 
                            path=paste0(getwd(),"/output/maxent_outputs4_cross"), 
                            args=prepPara(userfeatures="LQ",
                                          betamultiplier=1,
                                          doclamp = TRUE,
                                          projectionlayers="/Users/iel82user/Google Drive/1_osu_lab/projects/2017_7_workshop_enm_R/data/bioclim",
                                          replicates=5, ## 5 replicates
                                          replicatetype="crossvalidate") ) ##possible values are: crossvalidate,bootstrap,subsample


```

